---
title: "Clustering of samples individually"
output:
  html_document:
    toc: true
---

Notebook to visualise clustering of samples. Linear (PCA) and non-linear dimensional reduction techniques (UMAP/tSNE).

```{r load_modules, cache=FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(cache=FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(SingleCellExperiment)
library(ggplot2)
library(Seurat)

# Specify the rds files
sample_files <- str_replace(Sys.glob("RDS_objects.dir/*_filtered_clustered_SeuratObject.rds"), "RDS_objects.dir/", "")
# Sample names
sample_names <- gsub("_filtered_clustered_SeuratObject.rds", "", sample_files)
```

# Read in seurat objects

```{r read_seurat_objects, echo = FALSE}
for (i in sample_names){
  name <- paste0("RDS_objects.dir/", i, "_filtered_clustered_SeuratObject.rds")
  so <- readRDS(name)
  assign(paste("so", i, sep = "."), so)
}
```

# Variable features
Top variable features, number specified by user (default = 2000). Genes with highest cell-to-cell variation per sample (i.e. highly expressed in some cells, lowly expressed in others). The top 10 genes are labelled.

```{r variable_features, results='asis', echo = FALSE}

for (i in sample_names){
  so <- get(gsub("SAMPLE_FILE",i , "so.SAMPLE_FILE"))

  cat("## ",i,"\n")
  top10 <- head(VariableFeatures(so), 10)
  plot1 <- VariableFeaturePlot(so)
  plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)

  print(plot2)

  name<- paste0("Clustering_Figures.dir/VariableFeatures_", i, ".eps")
  postscript(name)
  print(plot2)
  dev.off()

  cat('\n\n')
}
```

# Dimensionality of the data

```{r dimensionality, results='asis', echo = FALSE}

for (i in sample_names){
  so <- get(gsub("SAMPLE_FILE",i , "so.SAMPLE_FILE"))

  cat("## ",i,"\n")

  jack_plot <- JackStrawPlot(so, dims = 1:15)

  elb_plt <- ElbowPlot(so)

  print(jack_plot)
  print(elb_plt)

  cat('\n\n')

}
```


# PCA- PC1 PC2 plot

```{r pca_dimplot, results='asis', echo = FALSE}

for (i in sample_names){
  so <- get(gsub("SAMPLE_FILE",i , "so.SAMPLE_FILE"))
  # Visualize PCA plot (PC1 and PC2)

  cat("## ",i,"\n")
  plt <- DimPlot(so, reduction = "pca")

  print(plt)

  name<- paste0("Clustering_Figures.dir/DimPlotPCA_", i, ".eps")
  postscript(name)
  print(plt)
  dev.off()

  cat('\n\n')
} 
```

# PCA Loading plot
Plot of top 30 genes contributing the most to PC1 and PC2.

```{r pca_loading, results='asis', echo = FALSE}

for (i in sample_names){
  so <- get(gsub("SAMPLE_FILE",i , "so.SAMPLE_FILE"))
  # Visualize PCA plot (PC1 and PC2)

  cat("## ",i,"\n")
  plt <- VizDimLoadings(so, dims = 1:2, reduction = "pca")

  print(plt)


  name<- paste0("Clustering_Figures.dir/LoadingPCA_", i, ".eps")
  postscript(name)
  print(plt)
  dev.off()

  cat('\n\n')

}
```

# PCA Dimension Heatmaps
Heatmap for 500 cells vs top genes for first 9 principle components

```{r pca_loading2, results='asis', echo = FALSE}

for (i in sample_names){
  so <- get(gsub("SAMPLE_FILE",i , "so.SAMPLE_FILE"))
  # Visualize PCA plot (PC1 and PC2)

  cat("## ",i,"\n")
  plt <- DimHeatmap(so, dims = 1:9, cells = 500, balanced = TRUE, reduction = "pca")


  print(plt)


  name<- paste0("Clustering_Figures.dir/DimensionHeatmap_", i, ".eps")
  postscript(name)
  print(plt)
  dev.off()

  cat('\n\n')
}
```


# t-SNE

```{r tsne_dimplot, results='asis', echo = FALSE}

for (i in sample_names){
  so <- get(gsub("SAMPLE_FILE",i , "so.SAMPLE_FILE"))

  cat("## ",i,"\n")
  plt <- DimPlot(so, reduction = "tsne")

  print(plt)

  name<- paste0("Clustering_Figures.dir/DimPlotTSNE_", i, ".eps")
  postscript(name)
  print(plt)
  dev.off()
  cat('\n\n')
}
```


# UMAP

```{r umap_dimplot, results='asis', echo = FALSE}

for (i in sample_names){
  so <- get(gsub("SAMPLE_FILE",i , "so.SAMPLE_FILE"))


  cat("## ",i,"\n")
  plt <- DimPlot(so, reduction = "umap")

  print(plt)

  name<- paste0("Clustering_Figures.dir/DimPlotUMAP_", i, ".eps")
  postscript(name)
  print(plt)
  dev.off()
  cat('\n\n')
}
```
