---
title: "Integrate data"
output: 
  html_document:
    code_folding: hide
---

The aim of this Rmardown document is to integrate all of the samples together in one seurat object. This Rmarkdown page will only render if there are more than 2 samples in the analysis.

# Import the Seurat objects

```{r}
knitr::opts_chunk$set(cache=FALSE, warning=FALSE,message=FALSE,error=FALSE)
library(Seurat)
library(ggplot2)
samples <- str_replace(Sys.glob("../kallisto.dir/*"), "../kallisto.dir/", "")

# Stop if only one file
if (length(samples) == 1){
  knitr::knit_exit()
}
```

# Import the data

```{r}
for (i in samples){
  name <- paste("../kallisto.dir/", i, "/seurat_output.rds", sep="")
  so <- readRDS(name)
  assign(paste("so", i, sep = "."), so)
  rm(so)
  
  }
```

# Integration Anchors

We then identify anchors using the FindIntegrationAnchors function, which takes a list of Seurat objects as input, and use these anchors to integrate the two datasets together with IntegrateData.

```{r}
data.list <- mget(ls(pattern = "^so"))

data.anchors <- FindIntegrationAnchors(object.list = data.list, dims = 1:15)

data.integrated <- IntegrateData(anchorset = data.anchors, dims = 1:15)
```

# Perform visualisation of the data


```{r, results='asis'}

# switch to integrated assay. The variable features of this assay are automatically
# set during IntegrateData
DefaultAssay(data.integrated) <- "integrated"

# Run the standard workflow for visualization and clustering
data.integrated <- ScaleData(data.integrated, verbose = FALSE)
data.integrated <- RunPCA(data.integrated, npcs = 30, verbose = FALSE)
data.integrated <- RunUMAP(data.integrated, reduction = "pca", dims = 1:30)
DimPlot(data.integrated, reduction = "umap")
```

# Cluster and plot UMAP

```{r}

data.integrated <- FindNeighbors(data.integrated, reduction = "pca", dims = 1:30)
data.integrated <- FindClusters(data.integrated, resolution = 0.35)
DimPlot(data.integrated, reduction = "umap", label = TRUE)

```

# Save normal data integration

```{r}
saveRDS(data.integrated, file = "integrated_CCA.rds")
```

# Harmony Analysis

We will next compare the typical CCA analysis of Seurat with Harmony 