---
title: "Annotation of cell types"
output:
  html_document:
    toc: true
---


Notebook to visualise cell type annotations. scClassify, singleR and clustifyr and Seurat techniques used.

```{r load_modules, cache=FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(cache=FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(ggplot2)
library(Seurat)

```

# scClassify
Annotation performed with scClasssify

## UMAP Annotation {.tabset .tabset-fade}

```{r umap_scc, results='asis', echo = FALSE}

plot_names <- Sys.glob("Annotation_Figures.dir/scclassify_UMAP*.eps")
if(length(plot_names) < 1){
  cat("No scClassify UMAP plots")
}else{

for (i in plot_names){
  name_obj <- gsub(".eps", "", gsub(pattern = "Annotation_Figures.dir/scclassify_UMAP_", "" , i))
  cat("### ",name_obj,"\n")
  knitr::include_graphics(path = i)
  cat('\n\n')
}
}
```

## tSNE Annotation {.tabset .tabset-fade}

```{r tsne_scc, results='asis', echo = FALSE}

plot_names <- Sys.glob("Annotation_Figures.dir/scclassify_tSNE*.eps")
if(length(plot_names) < 1){
  cat("No scClassify tSNE plots")
}else{

for (i in plot_names){
  name_obj <- gsub(".eps", "", gsub(pattern = "Annotation_Figures.dir/scclassify_tSNE_", "" , i))
  cat("### ",name_obj,"\n")
  knitr::include_graphics(path = i)
  cat('\n\n')
}
}
```

## Violin Plot Pre-defined Markers {.tabset .tabset-fade}

```{r violin_scc, results='asis', echo = FALSE}

plot_names <- Sys.glob("Annotation_Figures.dir/scclassify_ViolinPlot*.eps")
if(length(plot_names) < 1){
  cat("No scClassify violin plots, either because scClassify hasn't been run or because no predefined markers were supplied.")
}else{

for (i in plot_names){
  name_obj <- gsub(".eps", "", gsub(pattern = "Annotation_Figures.dir/scclassify_ViolinPlot_", "" , i))
  cat("### ",name_obj,"\n")
  knitr::include_graphics(path = i)
  cat('\n\n')
}
}
```

# Clustifyr
Annotation performed with Clustifyr

## UMAP Annotation {.tabset .tabset-fade}

```{r umap_clust, results='asis', echo = FALSE}

plot_names <- Sys.glob("Annotation_Figures.dir/clustifyr_UMAP*.eps")
if(length(plot_names) < 1){
  cat("No clustifyr UMAP plots")
}else{

for (i in plot_names){
  name_obj <- gsub(".eps", "", gsub(pattern = "Annotation_Figures.dir/clustifyr_UMAP_", "" , i))
  cat("### ",name_obj,"\n")
  knitr::include_graphics(path = i)
  cat('\n\n')
}
}
```

## tSNE Annotation {.tabset .tabset-fade}

```{r tsne_clust, results='asis', echo = FALSE}

plot_names <- Sys.glob("Annotation_Figures.dir/clustifyr_tSNE*.eps")
if(length(plot_names) < 1){
  cat("No clustifyr tSNE plots")
}else{

for (i in plot_names){
  name_obj <- gsub(".eps", "", gsub(pattern = "Annotation_Figures.dir/clustifyr_tSNE_", "" , i))
  cat("### ",name_obj,"\n")
  knitr::include_graphics(path = i)
  cat('\n\n')
}
}
```

## Violin Plot Pre-defined Markers {.tabset .tabset-fade}

```{r violin_clust, results='asis', echo = FALSE}

plot_names <- Sys.glob("Annotation_Figures.dir/clustifyr_ViolinPlot*.eps")
if(length(plot_names) < 1){
  cat("No clustifyr violin plots, either because clustifyr hasn't been run or because no predefined markers were supplied.")
}else{

for (i in plot_names){
  name_obj <- gsub(".eps", "", gsub(pattern = "Annotation_Figures.dir/clustifyr_ViolinPlot_", "" , i))
  cat("### ",name_obj,"\n")
  knitr::include_graphics(path = i)
  cat('\n\n')
}
}
```

# SingleR
Annotation performed with SingleR

## UMAP Annotation {.tabset .tabset-fade}

```{r umap_si, results='asis', echo = FALSE}

plot_names <- Sys.glob("Annotation_Figures.dir/singleR_UMAP*.eps")
if(length(plot_names) < 1){
  cat("No singleR UMAP plots")
}else{

for (i in plot_names){
  name_obj <- gsub(".eps", "", gsub(pattern = "Annotation_Figures.dir/singleR_UMAP_", "" , i))
  cat("### ",name_obj,"\n")
  knitr::include_graphics(path = i)
  cat('\n\n')
}
}
```

## tSNE Annotation {.tabset .tabset-fade}

```{r tsne_si, results='asis', echo = FALSE}

plot_names <- Sys.glob("Annotation_Figures.dir/singleR_tSNE*.eps")
if(length(plot_names) < 1){
  cat("No singleR tSNE plots")
}else{

for (i in plot_names){
  name_obj <- gsub(".eps", "",gsub(pattern = "Annotation_Figures.dir/singleR_tSNE_", "" , i))
  cat("### ",name_obj,"\n")
  knitr::include_graphics(path = i)
  cat('\n\n')
}
}
```

## Violin Plot Pre-defined Markers {.tabset .tabset-fade}

```{r violin_si, results='asis', echo = FALSE}

plot_names <- Sys.glob("Annotation_Figures.dir/singleR_ViolinPlot*.eps")
if(length(plot_names) < 1){
  cat("No singleR violin plots, either because clustifyr hasn't been run or because no predefined markers were supplied.")
}else{

for (i in plot_names){
  name_obj <- gsub(".eps", "", gsub(pattern = "Annotation_Figures.dir/singleR_ViolinPlot_", "" , i))
  cat("### ",name_obj,"\n")
  knitr::include_graphics(path = i)
  cat('\n\n')
}
}
```

## SingleR Misc {.tabset .tabset-fade}

```{r singleR_misc, results='asis', echo = FALSE}

plot_names1 <- Sys.glob("Annotation_Figures.dir/singleR_deltaDistribution*.eps")
plot_names2 <- Sys.glob("Annotation_Figures.dir/singleR_scoreDistribution*.eps")
plot_names3 <- Sys.glob("Annotation_Figures.dir/singleR_scoreHeatmap*.eps")
plot_names <- c(plot_names1,plot_names2,plot_names3)

if(length(plot_names) < 1){
  cat("No singleR misc plots.")
}else{

for (i in plot_names){
  name_obj <- gsub(".eps", "", gsub(pattern = "Annotation_Figures.dir/singleR_", "" , i))
  cat("### ",name_obj,"\n")
  knitr::include_graphics(path = i)
  cat('\n\n')
}
}
```

# Conserved Markers

## Dot Plots {.tabset .tabset-fade}

```{r dotplot, results='asis', echo = FALSE}

plot_names <- Sys.glob("Annotation_Figures.dir/DotPlot_TopConservedMarkers*.eps")

if(length(plot_names) < 1){
  cat("No dot plots. Has integrated_markers been run?")
}else{

for (i in plot_names){
  name_obj <- gsub(".eps", "", gsub(pattern = "Annotation_Figures.dir/DotPlot_TopConservedMarkers_", "" , i))
  cat("### ",name_obj,"\n")
  knitr::include_graphics(path = i)
  cat('\n\n')
}
}
```

## Top 9 Conserved markers per cluster {.tabset .tabset-fade}

```{r feature9, results='asis', echo = FALSE}

plot_names <- Sys.glob("Annotation_Figures.dir/FeaturePlot_TopConservedMarkers*.eps")

if(length(plot_names) < 1){
  cat("No feature plots. Has integrated_markers been run?")
}else{

for (i in plot_names){
  name_obj <- gsub(".eps", "", gsub(pattern = "Annotation_Figures.dir/FeaturePlot_TopConservedMarkers_", "" , i))
  cat("### ",name_obj,"\n")
  knitr::include_graphics(path = i)
  cat('\n\n')
}
}
```

# Clusters and markers
Pre-defined marker list general feature plot and violin plot (for each seurat cluster)

## Feature plot {.tabset .tabset-fade}

```{r feature9, results='asis', echo = FALSE}

plot_names <- Sys.glob("Annotation_Figures.dir/Clusters_FeaturePlot_*.eps")

if(length(plot_names) < 1){
  cat("No feature plots. Either integrated_markers has not been run or no pre-defined markers list given in yml.")
}else{

for (i in plot_names){
  name_obj <- gsub(".eps", "", gsub(pattern = "Annotation_Figures.dir/Clusters_FeaturePlot_", "" , i))
  cat("### ",name_obj,"\n")
  knitr::include_graphics(path = i)
  cat('\n\n')
}
}
```

## Violin plot {.tabset .tabset-fade}

```{r violin_clusters, results='asis', echo = FALSE}

plot_names <- Sys.glob("Annotation_Figures.dir/Clusters_ViolinPlot_*.eps")

if(length(plot_names) < 1){
  cat("No violin plots. Either integrated_markers has not been run or no pre-defined markers list given in yml.")
}else{

for (i in plot_names){
  name_obj <- gsub(".eps", "", gsub(pattern = "Annotation_Figures.dir/Clusters_ViolinPlot_", "" , i))
  cat("### ",name_obj,"\n")
  knitr::include_graphics(path = i)
  cat('\n\n')
}
}
```