---
title: "Clustering of samples individually"
output:
  html_document:
    toc: true
---

Notebook to visualise clustering of samples. Linear (PCA) and non-linear dimensional reduction techniques (UMAP/tSNE).

```{r load_modules, cache=FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(cache=FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(SingleCellExperiment)
library(Seurat)
library(biomaRt)

# Specify the rds files
sample_files <- str_replace(Sys.glob("RDS_objects.dir/*_filtered_clustered_SeuratObject.rds"), "RDS_objects.dir/", "")
# Sample names
sample_names <- gsub("_filtered_clustered_SeuratObject.rds", "", sample_files)
```

# Read in seurat objects

```{r read_seurat_objects, echo = FALSE}
for (i in sample_names){
  name <- paste0("RDS_objects.dir/", i, "_filtered_clustered_SeuratObject.rds")
  so <- readRDS(name)
  assign(paste("so", i, sep = "."), so)
}
```

# Variable features {.tabset .tabset-fade}
Top variable features, number specified by user (default = 2000). Genes with highest cell-to-cell variation per sample (i.e. highly expressed in some cells, lowly expressed in others). The top 10 genes are labelled.

First use biomart to get readable gene names from EnsemblIDs

```{r}
# useMart keeps not working every so often (issue https://github.com/grimbough/biomaRt/issues/31)
ensembl <- try(useMart("ensembl", dataset = "hsapiens_gene_ensembl"))
if(class(ensembl) == "try-error"){
  httr::set_config(httr::config(ssl_verifypeer = FALSE))
  ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
}

mapping <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"), mart = ensembl)


```


```{r variable_features, results='asis', echo = FALSE}

for (i in sample_names){
  so <- get(gsub("SAMPLE_FILE",i , "so.SAMPLE_FILE"))

  cat("## ",i,"\n")
  top10 <- head(VariableFeatures(so), 10)
  
  # Intersect top10 EnsemblIDs and gene names
  top10_genes <- top10 %>% as.data.frame() 
  colnames(top10_genes) <- "ensembl_gene_id"
  top10_genes <- left_join(top10_genes, mapping, na_matches = "na") 

  # plot variable features
  plot1 <- VariableFeaturePlot(so)
  
  # label with gene names
  plot2 <- LabelPoints(plot = plot1, points = top10, labels = top10_genes$hgnc_symbol, 
                       repel = TRUE, xnudge = 0, ynudge = 0)

  print(plot2)

  name<- paste0("Clustering_Figures.dir/VariableFeaturesGenes_", i, ".eps")
  postscript(name)
  print(plot2)
  dev.off()
  
  # label with ensemblIDs
  plot3 <- LabelPoints(plot = plot1, points = top10,  
                       repel = TRUE, xnudge = 0, ynudge = 0)

  print(plot3)

  name<- paste0("Clustering_Figures.dir/VariableFeaturesEnsemblID_", i, ".eps")
  postscript(name)
  print(plot3)
  dev.off()

  cat('\n\n')
}

```

# Dimensionality of the data {.tabset .tabset-fade}

```{r dimensionality, results='asis', echo = FALSE}

for (i in sample_names){
  so <- get(gsub("SAMPLE_FILE",i , "so.SAMPLE_FILE"))

  cat("## ",i,"\n")

  jack_plot <- JackStrawPlot(so, dims = 1:15)

  elb_plt <- ElbowPlot(so)

  print(jack_plot)
  print(elb_plt)
  
  name<- paste0("Clustering_Figures.dir/JackStrawPlot_", i, ".eps")
  postscript(name)
  print(jack_plot)
  dev.off()

  name<- paste0("Clustering_Figures.dir/ElbowPlot_", i, ".eps")
  postscript(name)
  print(elb_plt)
  dev.off()

cat('\n\n')

}
```


# PCA- PC1 PC2 plot {.tabset .tabset-fade}

```{r pca_dimplot, results='asis', echo = FALSE}

for (i in sample_names){
  so <- get(gsub("SAMPLE_FILE",i , "so.SAMPLE_FILE"))
  # Visualize PCA plot (PC1 and PC2)

  cat("## ",i,"\n")
  plt <- DimPlot(so, reduction = "pca")

  print(plt)

  name<- paste0("Clustering_Figures.dir/DimPlotPCA_", i, ".eps")
  postscript(name)
  print(plt)
  dev.off()

  cat('\n\n')
} 
```

# PCA Loading plot {.tabset .tabset-fade}
Plot of top 30 genes contributing the most to PC1 and PC2.

```{r pca_loading, results='asis', echo = FALSE}

for (i in sample_names){
  so <- get(gsub("SAMPLE_FILE",i , "so.SAMPLE_FILE"))
  # Visualize PCA plot (PC1 and PC2)

  cat("## ",i,"\n")
  plt <- VizDimLoadings(so, dims = 1:2, reduction = "pca")

  print(plt)


  name<- paste0("Clustering_Figures.dir/LoadingPCA_", i, ".eps")
  postscript(name)
  print(plt)
  dev.off()

  cat('\n\n')

}
```

# PCA Dimension Heatmaps {.tabset .tabset-fade}
Heatmap for 500 cells vs top genes for first 9 principle components

```{r pca_loading2, results='asis', echo = FALSE}

for (i in sample_names){
  so <- get(gsub("SAMPLE_FILE",i , "so.SAMPLE_FILE"))
  # Visualize PCA plot (PC1 and PC2)

  cat("## ",i,"\n")
  plt <- DimHeatmap(so, dims = 1:9, cells = 500, balanced = TRUE, reduction = "pca")


  print(plt)


  name<- paste0("Clustering_Figures.dir/DimensionHeatmap_", i, ".eps")
  postscript(name)
  print(plt)
  dev.off()

  cat('\n\n')
}
```


# t-SNE {.tabset .tabset-fade}

```{r tsne_dimplot, results='asis', echo = FALSE}

for (i in sample_names){
  so <- get(gsub("SAMPLE_FILE",i , "so.SAMPLE_FILE"))

  cat("## ",i,"\n")
  plt <- DimPlot(so, reduction = "tsne")

  print(plt)

  name<- paste0("Clustering_Figures.dir/DimPlotTSNE_", i, ".eps")
  postscript(name)
  print(plt)
  dev.off()
  cat('\n\n')
}
```


# UMAP {.tabset .tabset-fade}

```{r umap_dimplot, results='asis', echo = FALSE}

for (i in sample_names){
  so <- get(gsub("SAMPLE_FILE",i , "so.SAMPLE_FILE"))


  cat("## ",i,"\n")
  plt <- DimPlot(so, reduction = "umap")

  print(plt)

  name<- paste0("Clustering_Figures.dir/DimPlotUMAP_", i, ".eps")
  postscript(name)
  print(plt)
  dev.off()
  cat('\n\n')
}
```
